Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-05-09 manuscript.tex
Updates

\includegraphics[width=0.95\linewidth]{Images/Fig1_meanrtandhierplot_v2} \hfill{}

@@

\includegraphics[width=0.7\linewidth]{Images/Fig2_Hierarchical structure} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig4_distrtypefigure_v3} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig5_priorpredfigure_v2} \hfill{}

@@

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), \label{eq:chosenpriors1} \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions \texttt{brms} specifies on the standard deviations can be transformed to prior distributions on variances (\(var\)). Prior distributions on variances have a smaller scale than priors on standard deviation (\(sd\)). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \textcolor{blue}{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as ${sd}^2$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that match the prior distribution to the implied distribution.} In our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R\%20code/Prior\%20transformation}{Online Supplement}, we provide R-code with a numeric solution to the transformation. For the \texttt{brms} priors, we obtain the following implied prior distributions on variances:

@@

We will continue with step 3 of the prior specification process and check what the expected data would look like according to the selected priors. The results of the prior prediction for the symbolic distance effect with the priors from Equation \eqref{eq:chosenpriors1} with the number of participants \(n = 10\), the number of trials per condition \(k = 60\), and the number of simulations \(m = 1000\) are shown in panel A of Figure \ref{fig:priorpredfigurepanel}. Code for the simulations can be found in our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement}.

@@

The normal distribution on \(\mu_{\beta}\) and \(\mu_{\delta}\) seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's \emph{t}-distribution could be one of the reasons for the high range in variation. We change the prior for \(\mu_{\gamma}\) to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution.\footnote{Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution (Fink, 1997). For a normal likelihood, these are the normal distribution for the mean and the inverse gamma distribution for the variance.}\textcolor{blue}{However, in `rstan`, non-conjugate priors are commonly applied [e.g., @gelman2006prior]} Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks (e.g., Haaf \& Rouder, 2017), it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \ref{fig:priorpredfigurepanel}) and arrive at the following priors:\footnote{Because response times cannot be negative the prior for the overall baseline \(\mu_\gamma\) is a truncated normal with a lower bound of zero.}

@@

We follow the same procedure for prior specification in the log-normal model. As before, we prefer to apply conjugate priors. \textcolor{blue}{The conjugate priors that we applied to the normal model are also conjugate priors for the log-normal model [@fink1997compendium]. For instance, with a normal distribution as prior for the mean of the log-normal model, we assume that the mean of the parameter can be either negative, zero, or positive, and assign a higher probability to values closer to zero. Applying this conjugate prior, the posterior of the mean in the log-normal model will result in a normal distribution as well.}
As a starting point, we can use the adjusted priors of the normal model and perform a prior prediction, shown in panel C of Figure \ref{fig:priorpredfigurepanel}. These priors result in a slightly higher variance in RTs than we would expect, and, importantly, the mean response time differences are not centered around zero in all conditions. This pattern is inconsistent with our expectations, which means that prior distributions for the log-normal model need to be adjusted.

@@

There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used \texttt{R} packages, \texttt{rstan} (Stan Development Team, 2019a) and \texttt{brms} (Bürkner, 2017, 2018). We assume that readers are somewhat familiar with \texttt{R}. The package \texttt{rstan} provides the most modern algorithms and flexibility in model set-up. The package \texttt{brms} was created as an overlay to \texttt{rstan} and is based on the well-known \texttt{lme4} syntax (Bates, Mächler, Bolker, \& Walker, 2015). The goal of \texttt{brms} is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer a separate R Markdown file for each package in our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement} that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

\texttt{Rstan} allows the application of \texttt{Stan} (Carpenter et al., 2017), a probabilistic programming language, in \texttt{R}. The package uses a version of the Hamiltonian Monte Carlo (HMC) algorithm, No-U-turn sampler (NUTS; Hoffman \& Gelman, 2014), for sampling. \texttt{Rstan} requires two files: a \texttt{.stan} file containing the model specification and an R script containing the code for the model fitting. In the \texttt{.stan} file, the parameters, priors, and the likelihood are specified. An example \texttt{.stan} file of a simple, non-hierarchical normal model, where the variance is already known and the mean has to be estimated, is shown below (adapted example from Nicenboim, Schad, \& Vasishth, 2021, Chapter 10). The \texttt{.stan} file for the symbolic distance effect, including explanations, can be found in the \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/R\%20objects/rstan/normal\%20model/myModel.stan}{Online Supplement}. The file is divided into three sections: \texttt{data}, \texttt{parameters}, and \texttt{model}.

@@

\includegraphics[width=1\linewidth]{Images/Fig8_traceplots3pnew_v2} \hfill{}

@@

The model is fitted using the packages \texttt{rstan} and \texttt{brms}. \textcolor{blue}{We used different settings for `rstan` and `brms` due to the difference between the models for these packages. Although we tried to make the models as similar as possible, the models are not identical. This is partly due to difference in prior specification (i.e., specifying prior on the variances versus standard deviations).} For \texttt{rstan}, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For \texttt{brms}, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

\includegraphics[width=1\linewidth]{Images/Fig9_posteriormu_v3} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig10_randompar_v2} \hfill{}

@@

\includegraphics[width=0.75\linewidth]{Images/Fig11_modelestimatefigure2_onlyrstan_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig12_traceplots3plog_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig13_posteriormulog_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{Images/Fig14_randomparlog_v2} \hfill{}

@@

\includegraphics[width=0.75\linewidth]{Images/Fig15_inddiflogplot_onlyrstan_v2} \hfill{}

@@

The Bayes factor can also be interpreted as the relative evidence of two competing models. Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominatior in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers et al. (2018).

@@

\includegraphics[width=0.6\linewidth]{Images/Fig16_sdrplotside_v2} \hfill{}

@@

\includegraphics[width=0.7\linewidth]{Images/Fig17_normalbfplot01} \hfill{}

@@

\hypertarget{log-normal-model-3}{%
\subsubsection{Log-normal Model}\label{log-normal-model-3}}

We can take a similar approach for the log-normal model to compare the hierarchical models on varying complexity. We evaluate their performance in two scenarios that correspond to scenario 1 and 3 for the normal model. This means that in scenario 1 for the log-normal model, we estimated a null, digit, side and full model based on the prior specified in Equation, specifying large individual differences. For scenario 2, we remove individual variability from the side, digit, and full model. This resulted in the following posterior model probabilities shown in Table \ref{tab:postmodproblog}.

@@

The table shows that the digit model without individual digit effects from scenario 2 has the highest posterior model probability. This is in line with our findings for the normal model. Finally, we can compare the normal model with the highest marginal likelihood to the log-normal with the highest marginal likelihood. The normal digit model from scenario 3 to the log-normal model from scenario 2. The log marginal likelihood for the normal digit model without individual differences was \(6441.04\) while the log marginal likelihood for the log-normal digit model without individual differences was \(10637.45\). This indicates that the Bayes factor is in favor of the log-normal digit model over the normal digit model.

This example illustrates the benefits of comparing hierarchical models of varying complexity. While individual differences are a common phenomenon in psychology, there are instances where general effects are sufficient to describe the data adequately, or where only a random intercept, but no random slopes are necessary. As shown in Tables \ref{tab:postmodprob} and \ref{tab:postmodproblog}, Bayesian model comparisons can help researchers to identify the model that best describes the data.

@@

This tutorial showed the application of two R packages: \texttt{rstan} and \texttt{brms}. Which package to use depends mostly on the complexity of the model and your technical background. For the non-technical user, we recommend using \texttt{brms} because of its intuitive model specification. However, the default prior setup can be unsuitable for a specific research context and (virtually) always needs to be adapted. Additionally, users are limited to predefined parametrizations. \texttt{rstan} offers the most flexibility in model and prior specification. This flexibility comes with a cost: if the prior or the model is not well defined, results can be misleading.

@@

\hypertarget{section}{%
\subsection{\texorpdfstring{\textcolor{blue}{Technical Possibilities are Endless}}{}}\label{section}}

@@

\leavevmode\hypertarget{ref-Fink1997}{}%
Fink, D. (1997). A compendium of conjugate priors. https://www.johndcook.com/CompendiumOfConjugatePriors.pdf.

>>>>

\includegraphics[width=0.95\linewidth]{I - Images/Fig1_meanrtandhierplot_v2} \hfill{}

@@

\includegraphics[width=0.7\linewidth]{I - Images/Fig2_Hierarchical structure} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig4_distrtypefigure_v3} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig5_priorpredfigure_v2} \hfill{}

@@

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), \label{eq:chosenpriors1} \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions \texttt{brms} specifies on the standard deviations can be transformed to prior distributions on variances (\(var\)). Prior distributions on variances have a smaller scale than priors on standard deviation (\(sd\)). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \textcolor{blue}{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as ${sd}^2$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that match the prior distribution to the implied distribution.} In \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R\%20code/Prior\%20transformation}{Online Supplement J}, we provide R-code with a numeric solution to the transformation. For the \texttt{brms} priors, we obtain the following implied prior distributions on variances:

@@

We will continue with step 3 of the prior specification process and check what the expected data would look like according to the selected priors. The results of the prior prediction for the symbolic distance effect with the priors from Equation \eqref{eq:chosenpriors1} with the number of participants \(n = 10\), the number of trials per condition \(k = 60\), and the number of simulations \(m = 1000\) are shown in panel A of Figure \ref{fig:priorpredfigurepanel}. Code for the simulations can be found in \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/R\%20code/Figures/Fig5_priorpredictions_trunc_v3.R}{Online Supplement J}.

@@

The normal distribution on \(\mu_{\beta}\) and \(\mu_{\delta}\) seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's \emph{t}-distribution could be one of the reasons for the high range in variation. We change the prior for \(\mu_{\gamma}\) to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution.\footnote{Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution (Fink, 1997a). For a normal likelihood, these are the normal distribution for the mean and the inverse gamma distribution for the variance.} \footnote{\textcolor{blue}{However, in `rstan`, non-conjugate priors are commonly applied} (e.g., Gelman, 2006).} Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks (e.g., Haaf \& Rouder, 2017), it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \ref{fig:priorpredfigurepanel}) and arrive at the following priors:\footnote{Because response times cannot be negative the prior for the overall baseline \(\mu_\gamma\) is a truncated normal with a lower bound of zero.}

@@

We follow the same procedure for prior specification in the log-normal model. As before, we prefer to apply conjugate priors. \textcolor{blue}{The conjugate priors that we applied to the normal model are also conjugate priors for the log-normal model} (Fink, 1997b). \textcolor{blue}{For instance, with a normal distribution as prior for the mean of the log-normal model, we assume that the mean of the parameter can be either negative, zero, or positive, and assign a higher probability to values closer to zero. Applying this conjugate prior, the posterior of the mean in the log-normal model will result in a normal distribution as well.}
As a starting point, we can use the adjusted priors of the normal model and perform a prior prediction, shown in panel C of Figure \ref{fig:priorpredfigurepanel}. These priors result in a slightly higher variance in RTs than we would expect, and, importantly, the mean response time differences are not centered around zero in all conditions. This pattern is inconsistent with our expectations, which means that prior distributions for the log-normal model need to be adjusted.

@@

There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used \texttt{R} packages, \texttt{rstan} (Stan Development Team, 2019a) and \texttt{brms} (Bürkner, 2017, 2018). We assume that readers are somewhat familiar with \texttt{R}. The package \texttt{rstan} provides the most modern algorithms and flexibility in model set-up. The package \texttt{brms} was created as an overlay to \texttt{rstan} and is based on the well-known \texttt{lme4} syntax (Bates, Mächler, Bolker, \& Walker, 2015). The goal of \texttt{brms} is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer separate R Markdown files for each package in \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplements C-F} that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

\texttt{Rstan} allows the application of \texttt{Stan} (Carpenter et al., 2017), a probabilistic programming language, in \texttt{R}. The package uses a version of the Hamiltonian Monte Carlo (HMC) algorithm, No-U-turn sampler (NUTS; Hoffman \& Gelman, 2014), for sampling. \texttt{Rstan} requires two files: a \texttt{.stan} file containing the model specification and an R script containing the code for the model fitting. In the \texttt{.stan} file, the parameters, priors, and the likelihood are specified. An example \texttt{.stan} file of a simple, non-hierarchical normal model, where the variance is already known and the mean has to be estimated, is shown below (adapted example from Nicenboim, Schad, \& Vasishth, 2021, Chapter 10). The \texttt{.stan} file for the symbolic distance effect, including explanations, can be found in \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/blob/main/R\%20objects/rstan/normal\%20model/myModel.stan}{Online Supplement K}. The file is divided into three sections: \texttt{data}, \texttt{parameters}, and \texttt{model}.

@@

\includegraphics[width=1\linewidth]{I - Images/Fig8_traceplots3pnew_v2} \hfill{}

@@

The model is fitted using the packages \texttt{rstan} and \texttt{brms}. \textcolor{blue}{We used different settings for `rstan` and `brms` due to the difference between the models for these packages. Although we tried to make the models as similar as possible, the models are not identical. This is partly due to differences in prior specification (i.e., specifying prior on the variances versus standard deviations).} For \texttt{rstan}, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For \texttt{brms}, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

\includegraphics[width=1\linewidth]{I - Images/Fig9_posteriormu_v3} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig10_randompar_v2} \hfill{}

@@

\includegraphics[width=0.75\linewidth]{I - Images/Fig11_modelestimatefigure2_onlyrstan_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig12_traceplots3plog_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig13_posteriormulog_v2} \hfill{}

@@

\includegraphics[width=1\linewidth]{I - Images/Fig14_randomparlog_v2} \hfill{}

@@

\includegraphics[width=0.75\linewidth]{I - Images/Fig15_inddiflogplot_onlyrstan_v2} \hfill{}

@@

The Bayes factor can also be interpreted as the relative evidence of two competing models. Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominator in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers et al. (2018).

@@

\includegraphics[width=0.6\linewidth]{I - Images/Fig16_sdrplotside_v2} \hfill{}

@@

\includegraphics[width=0.7\linewidth]{I - Images/Fig17_normalbfplot01} \hfill{}

@@

\hypertarget{section}{%
\subsubsection{\texorpdfstring{\textcolor{blue}{Log-normal Model}}{}}\label{section}}

\textcolor{blue}{We can take a similar approach for the log-normal model to compare the hierarchical models on varying complexity. We evaluate their performance in two scenarios that correspond to scenario 1 and 3 for the normal model. This means that in scenario 1 for the log-normal model, we estimated a null, digit, side and full model based on the prior specified in Equation, specifying large individual differences. For scenario 2, we remove individual variability from the side, digit, and full model. This resulted in the following posterior model probabilities shown in Table} \ref{tab:postmodproblog}.

@@

\textcolor{blue}{The table shows that the digit model without individual digit effects from scenario 2 has the highest posterior model probability. This is in line with our findings for the normal model. Finally, we can compare the normal model with the highest marginal likelihood to the log-normal with the highest marginal likelihood. The normal digit model from scenario 3 to the log-normal model from scenario 2. The log marginal likelihood for the normal digit model without individual differences was $6441.04$ while the log marginal likelihood for the log-normal digit model without individual differences was $10637.45$. This indicates that the Bayes factor is in favor of the log-normal digit model over the normal digit model.}

The examples illustrate the benefits of comparing hierarchical models of varying complexity. While individual differences are a common phenomenon in psychology, there are instances where general effects are sufficient to describe the data adequately, or where only a random intercept, but no random slopes are necessary. As shown in Tables\} \ref{tab:postmodprob} and \ref{tab:postmodproblog}, Bayesian model comparisons can help researchers to identify the model that best describes the data.

@@

This tutorial showed the application of two R packages: \texttt{rstan} and \texttt{brms}. Which package to use depends mostly on the complexity of the model and your technical background. For the non-technical user, we recommend using \texttt{brms} because of its intuitive model specification. However, the default prior setup can be unsuitable for a specific research context and \textcolor{blue}{(virtually) always needs} to be adapted. Additionally, users are limited to predefined parametrizations. \texttt{rstan} offers the most flexibility in model and prior specification. This flexibility comes with a cost: if the prior or the model is not well defined, results can be misleading.

@@

\hypertarget{section-1}{%
\subsection{\texorpdfstring{\textcolor{blue}{Technical Possibilities are Endless}}{}}\label{section-1}}

@@

\leavevmode\hypertarget{ref-Fink1997}{}%
Fink, D. (1997a). A compendium of conjugate priors. https://www.johndcook.com/CompendiumOfConjugatePriors.pdf.

\leavevmode\hypertarget{ref-fink1997compendium}{}%
Fink, D. (1997b). A compendium of conjugate priors.
